### 2024-04-20 MSRI-16-1989-79.Rmd
starting 1

# Mixed Motives {#section-1}

:::: {.rmenv #chunk-1.1 title="1.1" latex="{1.1}"}
\oldpage{89}
For algebraic varieties, we have various parallel cohomology theories.
The most important for us will be de Rham and $\ell$-adic cohomology.

- **De Rham cohomology.**
  Let $k$ be a field of characteristic $0$, and $X$ an algebraic variety over $k$.
  Suppose that $X$ is smooth.
  The de Rham cohomology groups $\HH_\DR^i(X)$ are the hypercohomology groups of the de Rham complex:
  $$
    \HH_\DR^i(X)
    \coloneqq \bHH^i(X,\Omega_{X/k}^\bullet)
  $$
  cf. [@G].
  These are vector spaces over $k$.
  If $k'$ is an extension of $k$, and $X'$ over $k'$ is given by extension of scalars of $X$, then
  $$
    \HH_\DR^i(X') = \HH_\DR^i(X)\otimes_k k'.
  $$
  If $X$ is not smooth, then the de Rham complex no longer gives a reasonable theory.
  We can define the $\HH_\DR^i(X)$ by reduction to the smooth case, by the methods of [@D3], or, if $X$ admits an embedding into a smooth variety $Z$, as the hypercohomology of the de Rham complex of the formal completion of $Z$ along $X$ (R. Hartshorne, _On the de Rham cohomology of algebraic varieties_, Publ. Math. IHÉS **45** (1975), p. 5--99);
  more intrinsically, it is the crystalline cohomology of $X$ (A. Grothendieck, _Crystals and the de Rham cohomology of schemes_, Notes by J. Coates and O. Jussila, in: "dix exposés sur la cohomologie des schémas", North Holland (1968)).

- **$\ell$-adic cohomology.**
  Let $\ell$ be a prime number;
  if $k$ is an algebraically closed field of characteristic $\neq\ell$, then we have the $\ell$-adic theory $X\mapsto\HH^i(X,\QQ_\ell)$ that associates, to $X$ over $k$, cohomology groups which are vector spaces over $\QQ_\ell$ (cf. SGA 5, VI).
  They are defined from the cohomology groups with coefficients in $\ZZ/(\ell^n)$, and we allow ourselves to give, as reference for a theorem in $\ell$-adic cohomology, the place where its $\ZZ/(\ell^n)$ analogue is proved.
  The $\HH^i(X,\QQ_\ell)$ depend only on $X$.
  In particular, if $k$ is the algebraic closure of $k_0$, and if $X$ is given by extension of scalars of some $X_0$ over $k_0$, then $\Gal(k/k_0)$ acts (semi-$k$-linearly) on $X$, and thus acts on the $\HH^i(X,\QQ_\ell)$.
  This action is continuous.
  If $k'$ is an algebraically closed extension of $k$, and if $X'$ is given by extension of scalars of $X$, then $\HH^i(X,\QQ_\ell)\xrightarrow{\sim}\HH^i(X',\QQ_\ell)$.
  \oldpage{90}This follows by passing to the limit in the base change theorem for a smooth morphism [SGA 4, XVI, 1.2]: $k'$ is the filtrant inductive limit of the $k$-algebras $A$ with $\Spec(A)$ smooth over $k$.

If $k=\CC$, then we have the topological space $X(\CC)$ of points of $X$, as well as its rational cohomology $\HH^\bullet(X(\CC),\QQ)$.
We have canonical isomorphisms from [@G] and [SGA4, XVI, 4.1]:
$$
  \HH_\DR^i(X) = \HH^i(X(\CC),\QQ)\otimes_{\QQ}\CC
\tag{1.1.1}
$$
$$
  \HH^i(X,\QQ_\ell) = \HH^i(X(\CC),\QQ)\otimes_{\QQ}\QQ_\ell.
\tag{1.1.2}
$$

If $k$ is a field of characteristic $0$, and $\sigma\colon k\to\CC$ a complex embedding, with $\bar{k}$ the algebraic closure of $k$ in $\CC$ via $\sigma$, then we obtain the isomorphisms
$$
  \HH_\DR^i(X)\otimes_{k,\sigma}\CC = \HH^i(X(\CC),\QQ)\otimes_{\QQ}\CC
\tag{1.1.3}
$$
$$
  \HH^i(X\otimes\bar{k},\QQ_\ell) = \HH^i(X(\CC),\QQ)\otimes_{\QQ}\QQ_\ell
\tag{1.1.4}
$$
where $X(\CC)$ is the topological space of points of the complex algebraic variety given by the extension of scalars via $\sigma$ of $X$.

The existence of parallel cohomology theories lead A. Grothendieck to conjecture the existence, for all base fields $k$, of a motivic theory $X\mapsto\HH_\mot^i(X)$, defined on algebraic varieties (i.e. schemes of finite type) over $k$ and with values in a category $\mathcal{M}(k)$ (to be defined) of motives over $k$.
The known theories would then be deduced from the motivic theory by applying _realisation_ functors.

The category $\mathcal{M}(k)$ should be an abelian category, with $\Hom$ groups of finite dimension over $\QQ$.
It should be endowed with a tensor product $\otimes\colon\mathcal{M}(k)\times\mathcal{M}(k)\to\mathcal{M}(k)$ and associativity and commutative data $(X\otimes Y)\otimes Z\simto X\otimes(Y\otimes Z)$ and $X\otimes Y\to Y\otimes X$ satisfying the usual properties
--- more precisely, making $\mathcal{M}(k)$ into a Tannakian category [@Sa; @DM; @D4].
By the theory of Tannakian categories, $\mathcal{M}(k)$ would be the category of representations of a gerbe whose band is affine over $\Spec(\QQ)$.
For $k$ of characteristic $0$, the category $\mathcal{M}(k)$ with its tensor product should be equivalent to the category of representations of an scheme in affine groups (i.e. a proalgebraically affine group) over $\QQ$.

Each $\HH_\mot^i(X)$ would be a contravariant functor in $X$.
We should also have Künneth isomorphisms
$$
  \HH_\mot^i(X\times Y)
  \simeq \bigoplus_{i=j+k} \HH_\mot^j(X)\otimes\HH_\mot^k(Y)
\tag{1.1.5}
$$
\oldpage{91}giving rise to commutative diagrams
$$
  \begin{CD}
    \HH_\mot^i(X\times Y) @>\longleftarrow>> \HH_\mot^i(Y\times X)
  \\@VVV @VVV
  \\\HH_\mot^j(X)\otimes\HH_\mot^k(Y) @>>{(-i)^{jk}}> \HH_\mot^k(Y)\otimes\HH_\mot^j(X)
  \end{CD}
$$
$$
  \begin{CD}
    \HH_\mot^i((X\times Y)\times Z) @>\longleftarrow>> \HH_\mot^i(X\times(Y\times X))
  \\@VVV @VVV
  \\(\HH_\mot^j(X)\otimes\HH_\mot^k(Y))\otimes\HH_\mot^\ell(Z) @>>\longleftarrow> \HH_\mot^j(X)\otimes(\HH_\mot^k(Y)\otimes\HH_\mot^\ell(Z)).
  \end{CD}
$$

Each of the known cohomological theories should give rise to a "realisation" functor, compatible with the tensor product.
For example, for $k$ of characteristic $0$, we would have
$$
  \real_\DR\colon\mathcal{M}(k)
  \to \text{vector spaces over }k
$$
and, for $X$ an algebraic variety over $k$, a functorial isomorphism
$$
  \HH_\DR^i(X) = \real_\DR\HH_\mot^i(X)
$$
compatible with the Künneth isomorphisms.

The subcategory of $\mathcal{M}(k)$ _generated_ by a set $\mathcal{M}$ of motives is defined to be the smallest full subcategory of $\mathcal{M}(k)$ containing $\mathcal{M}$ that is stable under $\oplus$, $\otimes$, taking the dual, and sub-quotients.
If we only consider certain algebraic varieties $X$ over $k$, then it can be useful to consider, instead of $\mathcal{M}(k)$, the subcategory generated by the $\HH^i(X)$.
::::

@@

\oldpage{101}The motive $A$ is endowed with a product
$$
  A\otimes A\to A
\tag{1.16.1}
$$
namely the cup product, which makes $A$ a commutative ring with unit in the Tannakian category of motives (cf. [(5.3)](#chunk-5.3)).
On $A_\DR=E$, it is the product.
On $A_\BB$, it is given by $(q_1(\sigma))(q_2(\sigma))=(q_1(\sigma)q_2(\sigma))$.
::::

@@

The case $E=\Gm$ is particularly interesting: for every smooth scheme $S$ over $\Spec(\ZZ)$, we want, in the category of motives with integer coefficients over $S$,
$$
  \Gamma(S,\OO_S^\times) \simto \Ext^1(\ZZ(0),\ZZ(1)).
\tag{2.4.1}
$$
:::

@@

::: {.rmenv #chunk-3.1 title="3.1" latex="{3.1}"}
We denote by $P$ the torsor $P_{1,k}$ that we are going to define.
At the same time, we will define a trivialisation $u_\BB\in P_\BB$ of its Betti realisation.
Since $F^0\ZZ(k)_\DR=0$, we necessarily have that $P_\DR=\ZZ(k)_\DR=\QQ$.
The Betti structure is
$$
  \begin{aligned}
    P_\BB
    &= u_\BB + (2\pi i)^k\ZZ \subset \PP_{\DR,\CC} = \CC
    \quad\text{with}
  \\u_\BB
    &= -(k-1)!\zeta(k)
  \end{aligned}
\tag{3.1.1k}
$$
where $\zeta$ is the Riemann $\zeta$ function.

@@

Let $\alpha$ be an $N$-th root of $1$.
It defines $\alpha\colon\ZZ/N\to\ZZ/N(1)$ by $a\mapsto a\cdot\alpha$.
We thus obtain $\alpha^{\otimes(k-1)}\colon\ZZ/N(1)\to\ZZ/N(k)$.
Note that, for $\alpha=1$, this is the zero map.
We denote by $\alpha^{\otimes(k-1)}T(\alpha)$ the $\ZZ/N(k)$-torsor induced from $T(\alpha)$ by applying $\alpha^{\otimes(k-1)}$.
We have
$$
  P_{\ZZ/(N)}
  \coloneqq \sum_{\alpha^N=1} \alpha^{\otimes(k-1)}T_N(\alpha).
\tag{3.1.2k}
$$
\oldpage{113}It is equivalent to sum over the $\alpha\neq1$.
This definition still makes sense in any algebraic closure $\overline{k}$ of $\QQ$, and the action of $\Gal(\bar{\QQ}/\QQ)$ is given by transport of structure.

We now define the transition morphisms $P_{\ZZ/(NM)}\to P_{\ZZ/(N)}$.
We have
$$
  \prod_{u^M=z} (1-u)
  = 1-z
\tag{3.1.3}
$$
whence, for $z\neq0,1$, we obtain an isomorphism
$$
  T_N(z)
  = \prod_{u^M=z} T_N(u).
\tag{3.1.4}
$$
If $\beta^N=1$ and $\alpha^M=\beta$, then the reduction $\mod N$ of $\alpha\colon\ZZ/NM\to\ZZ/NM(1)$ is $\beta\colon\ZZ/N\to\ZZ/N(1)$, and that of $T_{NM}(\alpha)$ is $T_N(\alpha)$.
From this, we obtain an isomorphism
$$
  \begin{aligned}
    P_{\ZZ/(NM)}\mod N
    &= \sum_{\beta^N=1} \sum_{\alpha^M=\beta} \alpha^{\otimes(k-1)}T_{NM}(\alpha) \mod N
  \\&= \sum_{\beta^N=1} \beta^{\otimes(k-1)} \sum_{\alpha^M=\beta} T_N(\alpha)
  \\&= \sum_{\beta^N=1} \beta^{\otimes(k-1)} T_N(\beta)
  \\&= P_{\ZZ/(N)}.
  \end{aligned}
$$
This defines the desired transition morphism.

@@

If $z$ is not real and $\geqslant1$, then we have principal determinations
$$
  \prod_{u^M=z} (1-u)^{1/N}
  = (1-z)^{1/N}
\tag{3.1.5}
$$
as we verify by analytic continuation from $z=0$, taking (3.1.3) into account.
It follows from (3.1.5) that the $\comp_{\ZZ/(N),\BB}$ are compatible with the transition morphisms.
:::

::: {.rmenv #chunk-3.2 title="3.2" latex="{3.2}"}
\oldpage{114}Some preliminaries, before defining $\phi_p$.
The $k$-logarithm is defined for $|z|\leqslant1$ by
$$
  \ell_k(z)
  = \sum z^n/n^k.
\tag{3.2.1}
$$
The $\zeta(k)$ in the formula giving $P_\BB$ appears in our approach as the value at $1$ of $\ell_k$.
Let $p$ be a prime number and consider the formal series
$$
  \ell_k^{(p)}
  = \sum_{p\nmid n} z^n/n^k.
\tag{3.2.2}
$$
This is the Taylor series expansion of a rigid analytic function on the projective line over $\QQ_p$ with the open disc of radius $1$ centred at $1$ removed.
Indeed, we have
$$
  \ell_k^{(p)}(z)
  = \lim_N \frac{1}{1-z^{p^N}} \sum_{n=1, p\nmid n}^{p^N-1} z^n/n^k.
\tag{3.2.3}
$$
From (3.2.3) we get the identity
$$
  \ell_k^{(p)}(1/z)
  = -(-1)^k\ell_k^{(p)}(z).
\tag{3.2.4}
$$

A formal series calculation gives, for $(N,p)=1$,
$$
  N^{k-1}\sum_{u^N=z} \ell_k^{(p)}(u)
  = \ell_k^{(p)}(z).
\tag{3.2.5}
$$
This leads us to define the "principal value" at $1$
$$
  \vp\ell_k^{(p)}(1)
  \coloneqq \frac{N^{k-1}}{1-N^{k-1}} \sum_{\alpha^N=1,\alpha\neq1} \ell_k^{(p)}(\alpha).
\tag{3.2.6}
$$
This number is independent of the choice of $N$ coprime to $p$.
By (3.2.4), it is zero for $k$ even.

With this notation, the crystalline Frobenius $\phi_p$ is given by
$$
  \phi_p^{-1}(0)
  = (k-1)!p^k\cdot\vp\ell_k^{(p)}(1).
\tag{3.2.7k}
$$
:::

@@

With this notation, we have (cf. [@C; @Ko])
$$
  \vp \ell_k^{(p)}(1)
  = \zeta_p^{(p)}(k)
$$
and (3.2.7) can be rewritten as
$$
  \phi_p^{-1}(0)
  = (k-1)! p^k\zeta_p^{(p)}(k).
\tag{3.4.1k}
$$
:::

@@

Suppose that $F^0 M_\DR=0$.
Then $P_\DR=M_\DR$, and we can identify $P_\BB$ with a lateral class of $M_\BB\subset M_\DR\otimes\CC$.
Condition (a) then becomes $P_\BB\subset\frac{1}{n}M_\BB$.
We have $x=0$, and (b) is automatic.
We can write (d) as $\phi_p(0)=0$.
Let $p\in P_\BB$.
Denote by $p_M$ its image in $\frac{1}{n}M_\BB$, and set $p^\wedge=\comp_{\hZZ,\BB}(p)\in P_{\hZZ}$ and $(p_M)^\wedge=\comp_{\AAf,\BB}(p_M)\in M_\AAf$.
Condition (c) can then be written as
$$
  \sigma p^\wedge - p^\wedge
  = (\sigma-1)(p_M)^\wedge.
\tag{3.6.1}
$$

Let $M=\ZZ(k)$ with $k\geqslant1$.
Let $\sigma\in\Gal(\bar{\QQ}/\QQ)$ with image $c$ in $\hZZ^\times$, i.e. $\sigma(\zeta)=\zeta^c$ for $\zeta$ a root of unity.
Condition (3.6.1) can then be written as
$$
  \sigma p^\wedge - p^\wedge
  = (c^k-1)(p_M)^\wedge.
\tag{3.6.2}
$$
Modulo $N$, we obtain: let $c$ be an integer coprime to $Nn$, with $\sigma$ inducing $\zeta\mapsto\zeta^c$ on the $(Nn)$-th roots of unity.
Then $(c^k-1)p_M\in M_\BB$, and
$$
  \comp_{\hZZ,\BB}((c^k-1)p_M)
  \equiv \sigma p^\wedge - p^\wedge \mod{N}.
\tag{3.6.3}
$$

>>>>

@@

\oldpage{101}The motive $A$ is endowed with a product
$$
  A\otimes A\to A
\startverb\tag{1.16.1}\stopverb
$$
namely the cup product, which makes $A$ a commutative ring with unit in the Tannakian category of motives (cf. [(5.3)](#chunk-5.3)).
On $A_\DR=E$, it is the product.
On $A_\BB$, it is given by $(q_1(\sigma))(q_2(\sigma))=(q_1(\sigma)q_2(\sigma))$.
::::

@@

The case $E=\Gm$ is particularly interesting: for every smooth scheme $S$ over $\Spec(\ZZ)$, we want, in the category of motives with integer coefficients over $S$,
$$
  \Gamma(S,\OO_S^\times) \simto \Ext^1(\ZZ(0),\ZZ(1)).
\startverb\tag{2.4.1}\stopverb
$$
:::

@@

::: {.rmenv #chunk-3.1 title="3.1" latex="{3.1}"}
We denote by $P$ the torsor $P_{1,k}$ that we are going to define.
At the same time, we will define a trivialisation $u_\BB\in P_\BB$ of its Betti realisation.
Since $F^0\ZZ(k)_\DR=0$, we necessarily have that $P_\DR=\ZZ(k)_\DR=\QQ$.
The Betti structure is
$$
  \begin{aligned}
    P_\BB
    &= u_\BB + (2\pi i)^k\ZZ \subset \PP_{\DR,\CC} = \CC
    \quad\text{with}
  \\u_\BB
    &= -(k-1)!\zeta(k)
  \end{aligned}
\startverb\tag{3.1.1k}\stopverb
$$
where $\zeta$ is the Riemann $\zeta$ function.

@@

Let $\alpha$ be an $N$-th root of $1$.
It defines $\alpha\colon\ZZ/N\to\ZZ/N(1)$ by $a\mapsto a\cdot\alpha$.
We thus obtain $\alpha^{\otimes(k-1)}\colon\ZZ/N(1)\to\ZZ/N(k)$.
Note that, for $\alpha=1$, this is the zero map.
We denote by $\alpha^{\otimes(k-1)}T(\alpha)$ the $\ZZ/N(k)$-torsor induced from $T(\alpha)$ by applying $\alpha^{\otimes(k-1)}$.
We have
$$
  P_{\ZZ/(N)}
  \coloneqq \sum_{\alpha^N=1} \alpha^{\otimes(k-1)}T_N(\alpha).
\startverb\tag{3.1.2k}\stopverb
$$
\oldpage{113}It is equivalent to sum over the $\alpha\neq1$.
This definition still makes sense in any algebraic closure $\overline{k}$ of $\QQ$, and the action of $\Gal(\bar{\QQ}/\QQ)$ is given by transport of structure.

We now define the transition morphisms $P_{\ZZ/(NM)}\to P_{\ZZ/(N)}$.
We have
$$
  \prod_{u^M=z} (1-u)
  = 1-z
\startverb\tag{3.1.3}\stopverb
$$
whence, for $z\neq0,1$, we obtain an isomorphism
$$
  T_N(z)
  = \prod_{u^M=z} T_N(u).
\startverb\tag{3.1.4}\stopverb
$$
If $\beta^N=1$ and $\alpha^M=\beta$, then the reduction $\mod N$ of $\alpha\colon\ZZ/NM\to\ZZ/NM(1)$ is $\beta\colon\ZZ/N\to\ZZ/N(1)$, and that of $T_{NM}(\alpha)$ is $T_N(\alpha)$.
From this, we obtain an isomorphism
$$
  \begin{aligned}
    P_{\ZZ/(NM)}\mod N
    &= \sum_{\beta^N=1} \sum_{\alpha^M=\beta} \alpha^{\otimes(k-1)}T_{NM}(\alpha) \mod N
  \\&= \sum_{\beta^N=1} \beta^{\otimes(k-1)} \sum_{\alpha^M=\beta} T_N(\alpha)
  \\&= \sum_{\beta^N=1} \beta^{\otimes(k-1)} T_N(\beta)
  \\&= P_{\ZZ/(N)}.
  \end{aligned}
$$
This defines the desired transition morphism.

@@

If $z$ is not real and $\geqslant1$, then we have principal determinations
$$
  \prod_{u^M=z} (1-u)^{1/N}
  = (1-z)^{1/N}
\startverb\tag{3.1.5}\stopverb
$$
as we verify by analytic continuation from $z=0$, taking (3.1.3) into account.
It follows from (3.1.5) that the $\comp_{\ZZ/(N),\BB}$ are compatible with the transition morphisms.
:::

::: {.rmenv #chunk-3.2 title="3.2" latex="{3.2}"}
\oldpage{114}Some preliminaries, before defining $\phi_p$.
The $k$-logarithm is defined for $|z|\leqslant1$ by
$$
  \ell_k(z)
  = \sum z^n/n^k.
\startverb\tag{3.2.1}\stopverb
$$
The $\zeta(k)$ in the formula giving $P_\BB$ appears in our approach as the value at $1$ of $\ell_k$.
Let $p$ be a prime number and consider the formal series
$$
  \ell_k^{(p)}
  = \sum_{p\nmid n} z^n/n^k.
\startverb\tag{3.2.2}\stopverb
$$
This is the Taylor series expansion of a rigid analytic function on the projective line over $\QQ_p$ with the open disc of radius $1$ centred at $1$ removed.
Indeed, we have
$$
  \ell_k^{(p)}(z)
  = \lim_N \frac{1}{1-z^{p^N}} \sum_{n=1, p\nmid n}^{p^N-1} z^n/n^k.
\startverb\tag{3.2.3}\stopverb
$$
From (3.2.3) we get the identity
$$
  \ell_k^{(p)}(1/z)
  = -(-1)^k\ell_k^{(p)}(z).
\startverb\tag{3.2.4}\stopverb
$$

A formal series calculation gives, for $(N,p)=1$,
$$
  N^{k-1}\sum_{u^N=z} \ell_k^{(p)}(u)
  = \ell_k^{(p)}(z).
\startverb\tag{3.2.5}\stopverb
$$
This leads us to define the "principal value" at $1$
$$
  \vp\ell_k^{(p)}(1)
  \coloneqq \frac{N^{k-1}}{1-N^{k-1}} \sum_{\alpha^N=1,\alpha\neq1} \ell_k^{(p)}(\alpha).
\startverb\tag{3.2.6}\stopverb
$$
This number is independent of the choice of $N$ coprime to $p$.
By (3.2.4), it is zero for $k$ even.

With this notation, the crystalline Frobenius $\phi_p$ is given by
$$
  \phi_p^{-1}(0)
  = (k-1)!p^k\cdot\vp\ell_k^{(p)}(1).
\startverb\tag{3.2.7k}\stopverb
$$
:::

@@

With this notation, we have (cf. [@C; @Ko])
$$
  \vp \ell_k^{(p)}(1)
  = \zeta_p^{(p)}(k)
$$
and (3.2.7) can be rewritten as
$$
  \phi_p^{-1}(0)
  = (k-1)! p^k\zeta_p^{(p)}(k).
\startverb\tag{3.4.1k}\stopverb
$$
:::

@@

Suppose that $F^0 M_\DR=0$.
Then $P_\DR=M_\DR$, and we can identify $P_\BB$ with a lateral class of $M_\BB\subset M_\DR\otimes\CC$.
Condition (a) then becomes $P_\BB\subset\frac{1}{n}M_\BB$.
We have $x=0$, and (b) is automatic.
We can write (d) as $\phi_p(0)=0$.
Let $p\in P_\BB$.
Denote by $p_M$ its image in $\frac{1}{n}M_\BB$, and set $p^\wedge=\comp_{\hZZ,\BB}(p)\in P_{\hZZ}$ and $(p_M)^\wedge=\comp_{\AAf,\BB}(p_M)\in M_\AAf$.
Condition (c) can then be written as
$$
  \sigma p^\wedge - p^\wedge
  = (\sigma-1)(p_M)^\wedge.
\startverb\tag{3.6.1}\stopverb
$$

Let $M=\ZZ(k)$ with $k\geqslant1$.
Let $\sigma\in\Gal(\bar{\QQ}/\QQ)$ with image $c$ in $\hZZ^\times$, i.e. $\sigma(\zeta)=\zeta^c$ for $\zeta$ a root of unity.
Condition (3.6.1) can then be written as
$$
  \sigma p^\wedge - p^\wedge
  = (c^k-1)(p_M)^\wedge.
\startverb\tag{3.6.2}\stopverb
$$
Modulo $N$, we obtain: let $c$ be an integer coprime to $Nn$, with $\sigma$ inducing $\zeta\mapsto\zeta^c$ on the $(Nn)$-th roots of unity.
Then $(c^k-1)p_M\in M_\BB$, and
$$
  \comp_{\hZZ,\BB}((c^k-1)p_M)
  \equiv \sigma p^\wedge - p^\wedge \mod{N}.
\startverb\tag{3.6.3}\stopverb
$$
